How is experience transformed into memory?

### 2018-08-27 paper/main.tex
added back brain analyses and added paxtons control analyses and text

\title{How is experience transformed into memory?}
\author{Andrew C. Heusser and Jeremy R. Manning\\Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NH 03755, USA\\Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\begin{figure}[th!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic_blur.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participant's verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in green) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

\subsubsection{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text, see Methods for analysis details). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable.

After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}b. Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a timepoint (68 to 294) by timepoints by timepoints (range: 68 to 294) correlation matrix. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d. Like the video model, each participant's recall correlation matrix exhibited a strong block-diagonal structure (Supp. Fig.~\ref{fig:corrmats}). Notably, this suggests that participants recounted the video in discriminable segments, likely related to the recall of specific events from the video.

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e., topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} A). Group-averaged correlation between each video event and closest recall event (per subject). Error bars represent 95\% confidence intervals. B). Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g.\ the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. C). Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are all individual recall events across all participants, where the colors refer to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

@@

Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words. Our work characterizes experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape.

In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g.\ percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

@@

To conclude, we'll revisit the question of what it means to remember something. Our view is that ``successful remembering'' is about recovering the ``trajectory'' of an experience, rather than the ability to recognize/recall any of its particular isolated details. Decades of research suggest that episodic memories are not veridical and context-free snapshots of the past, and so treating (and modelling) them as such is overly simplistic at best (and counterproductive at worst). While its undeniable that these models have been useful in advancing our understanding of human memory, they are severely limited in their ability to explain memory for naturalistic experiences. Real-life experiences are highly structured in time, and so to have a complete understanding of the human memory system, our experiments and models of memory should not ignore this fact. This work provides a methodological and theoretical advance in our understanding of what it means to remember.

@@

@@

@@

\section*{Supplemental Methods}
\subsection{List-learning analyses}
\textbf{Overall Accuracy.}  To get an overall measure of the quantity of information recalled, we computed the proportion of sucessfully recalled events by counting the number of unique recall events identified by the HMM model and dividing by the total number of video events.  We performed this analysis for each participant separately.

>>>>

\title{How is experience transformed into memory?}
% \title{Capturing the transformation of experience into memory}
\author{Andrew C. Heusser and Jeremy R. Manning\\Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\begin{figure}[th!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic_bw.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participant's verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in green) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

\subsubsection{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.~\ref{fig:schematic} for example text). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic ``state'' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable. We found that the ``narrative details'' (e.g. a few sentences summarizing the specific happenings of the scene) of the video were the most important feature driving this model structure (see Supp. Fig.~\ref{fig:feature-importance}a).

After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}b. Notably, participant \#13's recall model appears visually similar to the video model. Similar to the video model analysis (Supp. Fig.~\ref{fig:feature-importance}a), We found that the ``narrative details'' feature was most important for driving the video/recall relationship (Supp. Fig.~\ref{fig:feature-importance}b).  Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a timepoint (68 to 294) by timepoints by timepoints (range: 68 to 294) correlation matrix. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d. Like the video model, each participant's recall correlation matrix exhibited a strong block-diagonal structure (Supp. Fig.~\ref{fig:corrmats}). Notably, this suggests that participants recounted the video in discriminable segments, likely related to the recall of specific events from the video.

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e., topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

@@

\subsubsection{Brain regions that mirror the correlational structure of naturalistic stimuli and memories}.
\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_brains_panels.pdf}
\caption{\small \textbf{Video and recall model correlational structure searchlight analysis.} A) The correlation matrix representing the temporal structure of the video model. Below it is a surface plot of brain regions whose activity patterns mirrored the correlational structure of video model during viewing.  B) The left matrix is S13's video-recall correlation matrix before warping. The yellow stars represent the path recovered by dynamic time warping. The right matrix is the result after warping. The same warping path was applied to the fMRI timeseries (separately for each participant). A surface plot of brain regions whose activity patterns mirrored the participant-specific recall correlation matrix. Significance threshold (p$<$.05) for both video and recall analyses derived from a permutation procedure by phase shifting the video model and recomputing the z-scores.}
\label{fig:brainz}
\end{figure}

Previous work suggests that the brain may represent our experiences as ``events'' with different time scales in a nested hierarchy ~\citep{LernEtal11, ChenEtal17, BaldEtal17}. In this next analysis we asked: what brain regions mirror the event-like correlational structure of our topic-based video model during viewing? We used a searchlight procedure (see Methods for details) to identify regions whose timepoint-by-timepoint correlational structure matched the correlational structure of our model. This analysis revealed a network of regions including bilateral frontal and cingulate cortex (Fig. ~\ref{fig:brainz}). Thus, these regions may play a role in maintaining information relevant to the narrative trajectory of the video.

While each participant remembered roughly the same sequence of events, there were subtle idiosyncracies in the way each individual described the events. For example, the event by event similarity structure (e.g. the off-diagonal values of the recall correlation matrices) differed across individuals (Supp. Fig.~\ref{fig:corrmats}). We hypothesized that idiosyncracies in the recall dynamics across subjects might be related to the temporal structure of brain networks are representing the events during video viewing. In this analysis, we identified brain areas whose responses at the time of viewing exhibited the same (segmented) idiosyncratic dynamics that later appeared in that participants’ recalls. Using a procedure similar to the approach above (see Methods for details), we searched for regions whose correlational structure during viewing was related to the idiosyncratic structure of the participant's recall model. The resulting map suggests that the correlational structure of a network of brain regions including the ventromedial prefrontal cortex, anterior cingulate, and right medial temporal lobe mirror the idiosyncratic structure of participants'' recalls (Fig. ~\ref{fig:brainz}).

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} A). Group-averaged correlation between each video event and closest recall event. Error bars represent 95\% confidence intervals. Stars indicate top (green) and bottom (red) 3 remembered events across participants B). Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g.\ the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. C). Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are all individual recall events across all participants, where the colors refer to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

@@

Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words. Our work characterizes experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape. In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g.\ percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

Prior work on neural responses during naturalistic experiences and recall has largely focused on identifying brain regions whose responses are reliably similar across individuals ~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17}. This allows one to identify which regions might be processing or representing the stimulus, or retrieving details of the experience during recall. Our approach is fundamentally different. We ask: which brain regions during movie viewing match the movie structure and how individual participants will recall the movie later (in terms of the temporal correlations of their recall topic trajectory)? This latter approach highlights regions such as the medial prefrontal cortex (mPFC) and medial temporal lobes (MTL) that may respond in idiosyncratic ways across individuals, but that nonetheless play an important role in encoding experience into memory. In other work, the mPFC has been suggested to play a role in the representation of a ``schema'', or a network prior knowledge that is consistent with the current task ~\citep{KestEtal12}. One interpretation for mPFC's involvement in our work is that mPFC represents participant-specific relationships with the content of the video, and its representation during scene viewing is related to how a participant later describes the scene. MTL activity patterns also matched participant's recall, which is consistent with a large body of literature highlighting the role of the MTL in episodic memory encoding ~\citep{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.

@@

To conclude, we'll revisit the question of what it means to remember something. Our view is that ``successful remembering'' is about recovering the ``trajectory'' of an experience, rather than the ability to recognize/recall any of its particular isolated details. Decades of research suggest that episodic memories are not veridical and context-free snapshots of the past, and so treating (and modelling) them as such is overly simplistic at best (and counterproductive at worst). While its undeniable that these models have been useful in advancing our understanding of human memory, they are severely limited in their ability to explain memory for naturalistic experiences. Real-life experiences are highly structured in time, and so to have a complete understanding of the human memory system, our experiments and models of memory should not ignore this fact. Our work provides a theoretical advance in our understanding of what it means to remember as well as a novel methodological tool to study it.

@@

\subsection{fMRI analyses}
Seventeen participants watched the first 50 minutes of Episode 1 of BBC's Sherlock. The video was split into two parts of approximately equal length (946 and 1030 TRs). All data were preprocessed and transformed to 3mm MNI space as described in ~\citep{ChenEtal17}. Data were z-scored across time at every voxel. 6mm smoothing was applied.
Files are cropped so that all video-viewing data are aligned across participants, and all recall data are aligned to the scene timestamps below. The cropping includes a constant 3-TR (4.5 sec) shift to correct for hemodynamic lag.

\subsubsection{Searchlight analysis}
Our multivariate analyses were designed to capture brain regions whose timepoint-by-timepoint correlational structure mirrors the correlational structure of the video model as well as participant-specific recall topic models during video viewing. We conducted a searchlight analysis (5x5x5 voxel cube) where for each cube, we correlated the model timepoint-by-timepoint correlation matrix with the neural correlation matrix. To aggregate across participants, we Fisher's z-transformed the correlations and then averaged.  To assess significance, we recomputed this group analysis 100 times, but randomly phase shifted the model (by \# of timepoints - 1) by the same amount for each participant but different amounts for each permutation to build a null distribution of correlation values. Finally, we thresholded the group averaged correlation maps where the `real' correlation value for a given voxel exceeded the 95th percentile of the null distribution. To correct for non-linearities between the viewing time and recall time, for each participant we used dynamic time warping to temporally align the matrices. The algorithm recovers a path of coordinates that would bring the video and recall model in maximal temporal alignment. We used this path to warp the fMRI data and the recall model into temporal alignment (separately for each participant).

@@

\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp4_feature_value.pdf}
\caption{\small \textbf{Impact of individual features on topic modeling analysis.} Contribution of each feature to model structure. Bars represent the correlation of a video model trained in the absence of a given feature to the model trained on all features. B).Contribution of each feature to video model/recall model relationship. The leftmost bar represents the across-subjects mean correlation between the video model and recall models trained on all features. Subsequent bars represent the same relationship between video and recall models trained in the absence of a given feature. Error bars are the standard error of the mean across subjects.}
\label{fig:feature-importance}
\end{suppfigure}

@@

\section*{Supplemental Methods}
\subsection{Quantifying the importance of features}
To determine the contribution of each feature to the structure of the video model, we examined the similarity between the temporal structure of models trained in the absence of a single feature and that of our original (i.e. full) model. First, we iteratively removed one transcribed feature from the scene descriptions and constructed a timepoints (1976) by topis (100) matrix using a topic model fit to the remaining features. We then represented the original model as well as the new model’s temporal structure as a timepoints (1976) by timepoints (1976) correlation matrix. Finally, we vectorized these correlation matrices and correlated them to each other resulting in a single number (for each feature removed from the model) representing the similarity of the ``feature-removed'' models to the full model.

In order to ascertain which features were important in relating the recall models to the video model, we similarly compared to the temporal structures of video and recall models deprived of a single feature at a time. For each feature removed, we transformed each subject’s recall transcript using a model trained on the feature-deprived video text windows (of 50 scene segments), and resampled the recall timeseries to match the shape of the movie model (1976 timepoints). We then represented the temporal structures of each subject’s recall model as timepoints (1976) by timepoints (1976) correlation matrices and computed the average correlation with the temporal structure of the video model (across subjects).

\subsection{List-learning analyses}
\textbf{Overall Accuracy.}  To get an overall measure of the quantity of information recalled, we computed the proportion of sucessfully recalled events by counting the number of unique recall events identified by the HMM model and dividing by the total number of video events.  We performed this analysis for each participant separately.
